------------------------------------------------------------------------

``` yaml
---
title: "Nearest Neighbor Propensity Score Matching with MatchIt"
author: "Your Name"
date: "2025-01-26"
output:
  html_document:
    toc: true
    toc_depth: 3
---
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)
```

# Propensity Score Matching with **MatchIt**

> **Research Question**:\
> What is the causal effect between smokers with a *high cholesterol* level and smokers with a *normal cholesterol* level on death by 1992?

```{r}
load("C:/Users/P094412/Documents/bcf-linear/nhefs_prepared_day2.rdata")
```

## (a) Interpretation of ACET

```{r a-interpretation, eval=FALSE}
# Interpretation:
# The ACET is the average effect for those with high cholesterol, 
# interpreted as how their risk of death would change if they had normal cholesterol instead.
```

## (b) 1:1 Nearest Neighbor Matching *without replacement*

```{r b-match-no-replace}
# If needed, install MatchIt:
# install.packages("MatchIt")

library(MatchIt)

# Define the propensity score formula
eq_psmod <- Cholesterol_dicho ~ active + alcoholfreq + asthma + bronch + 
  exercise + income + race + sex + bmi + age

# Perform 1:1 nearest neighbor matching without replacement
match_no_replace <- matchit(formula = eq_psmod,
                            data = dat, 
                            method = "nearest",
                            distance = "glm")

# Check how many unmatched
summary(match_no_replace)$nn
```

## (c) Properties of the Matching Model

```{r c-model-props, eval=FALSE}
# This is 1:1 NN matching in descending order of PS, without replacement, no caliper.
```

## (d) Model Used Internally

```{r d-internal-model, eval=FALSE}
# matchit(distance = "glm") -> uses a logistic regression on all covariates in eq_psmod.
```

## (e) Estimate PS Externally and Pass to `matchit`

```{r e-external-ps}
# Fit the logistic regression externally
ps.mod <- glm(eq_psmod, data = dat, family = binomial)

# Get propensity score
dat$ps <- predict(ps.mod, newdata = dat, type = "response")

# Perform matching by passing the pre-computed distance
match_external_ps <- matchit(formula = eq_psmod,
                             data = dat,
                             method = "nearest",
                             distance = dat$ps)

summary(match_external_ps)$nn
```

## (f) Balance Check on Matched Sample

```{r f-balance-check}
# Quick check of balance
plot(summary(match_no_replace))
```

> Because only 18 (or so) treated units are unmatched, balance may not improve much.

## (g) Improving Balance with a Caliper (0.2 and 0.05)

```{r g-calipers}
match_cal20 <- matchit(eq_psmod, data = dat,
                       method = "nearest",
                       distance = "glm",
                       caliper = 0.2)
match_cal05 <- matchit(eq_psmod, data = dat,
                       method = "nearest",
                       distance = "glm",
                       caliper = 0.05)

# Number of matched
summary(match_cal20)$nn
summary(match_cal05)$nn
```

## (h) 1:1 Matching -\> Equal \# Treated and Controls

```{r h-one-one, eval=FALSE}
# Because ratio = 1 by default in matchit, we get the same # matched in T and C.
```

## (i) Fewer Matches with Smaller Caliper

```{r i-explanation, eval=FALSE}
# A stricter (smaller) caliper excludes more treated who don't find a close control match.
```

## (j) Love Plots via **cobalt**

```{r j-love-plots, eval=FALSE}
# install.packages("cobalt") if needed
library(cobalt)

love.plot(match_cal20, 
          stats = "mean.diffs", 
          abs = TRUE,
          var.order = "unadjusted",
          threshold = 0.1)

love.plot(match_cal05, 
          stats = "mean.diffs", 
          abs = TRUE,
          var.order = "unadjusted",
          threshold = 0.1)
```

## (k) Consequences of Excluding Unmatched Treated

```{r k-exclusion, eval=FALSE}
# We no longer estimate ACET for all treated, only those who found a match.
# This can reduce external validity if many are excluded.
```

## (l) Check Caliper Values

```{r l-check-caliper}
match_cal20$caliper
match_cal05$caliper

sd(match_cal20$distance) * 0.2
sd(match_cal05$distance) * 0.05
```

## (m) Inspecting Variance Ratios

```{r m-variance-ratios, eval=FALSE}
summary(match_cal20)   # shows var. ratio, SMD, etc.
```

## (n) Matching with Replacement

```{r n-with-replacement}
match_cal20r <- matchit(eq_psmod, data = dat,
                        method = "nearest",
                        distance = "glm",
                        caliper = 0.2,
                        replace = TRUE)

match_cal05r <- matchit(eq_psmod, data = dat,
                        method = "nearest",
                        distance = "glm",
                        caliper = 0.05,
                        replace = TRUE)

summary(match_cal20r)$nn
summary(match_cal05r)$nn
```

## (o) More Treated Units Matched

```{r o-advantage, eval=FALSE}
# Because with replacement, we can re-use the same control multiple times 
# and thus keep most/all treated matched.
```

## (p) Fewer Unique Controls

```{r p-why-fewer-controls, eval=FALSE}
# The best (closest) controls get matched repeatedly.
```

## (q) Adding Quadratic Terms to Fix Variance Ratios

```{r q-bmi-age2}
eq_psmod2 <- Cholesterol_dicho ~ active + alcoholfreq + asthma + bronch + exercise +
  income + race + sex + poly(age, 2) + poly(bmi, 2)

match_cal20r_v2 <- matchit(eq_psmod2, data = dat,
                           method = "nearest",
                           distance = "glm",
                           caliper = 0.2,
                           replace = TRUE)

summary(match_cal20r_v2)
```

## (r) Jitter Plot Without Replacement

```{r r-jitter-plot-norep, eval=FALSE}
plot(match_cal20, type = "jitter", interactive = FALSE)
```

## (s) Jitter Plot With Replacement

```{r s-jitter-plot-rep, eval=FALSE}
plot(match_cal20r, type = "jitter", interactive = FALSE)
```

## (t) Extract Matched Data (`get_matches`) -\> All Weights = 1

```{r t-get-matches}
dat_matched <- get_matches(match_cal20r)
head(dat_matched)
table(dat_matched$weights)
```

## (u) How Often is the Same Control Matched?

```{r u-control-usage}
table(table(subset(dat_matched, Cholesterol_dicho == 0)$id))
```

## Estimate the Causal (Log) Odds Ratio

```{r final-estimation}
# For cluster-robust SE:
library(lmtest)
library(sandwich)

mod <- glm(death ~ Cholesterol_dicho,
           data = dat_matched,
           family = binomial)

# Adjust SE for clustering by subclass + id
coeftest(mod, vcov. = vcovCL, cluster = ~ subclass + id)
```

> This gives us a **cluster-robust** standard error, accounting for matched sets and repeated usage of some controls.

------------------------------------------------------------------------

**End of R Markdown**

**Instructions**:

1.  Copy all of the text above (the YAML header, code, and text) into a file named something like `matching_tutorial.Rmd`.
2.  Open it in RStudio (or any other environment that supports R Markdown).
3.  Click **Knit** to produce an HTML document.
    -   Make sure you have the necessary libraries installed:

        ``` r
        install.packages(c("MatchIt", "cobalt", "lmtest", "sandwich"))
        ```
4.  Follow the rendered document to see the code output and explanations.
