---
title: "Causal Inference and Propensity Score Methods: R Tutorial Day 4 Solutions"
author: "Adapted from Thomas Klausch"
date: "`r Sys.Date()`"
output: html_document
---

## Instructions

1.  **Copy all text from this file** into a new file named, for example, `day4_solutions.Rmd`.
2.  **Place** your `connors.rdata` file in the same directory (or update the path in the code).
3.  Open `day4_solutions.Rmd` in RStudio and click **Knit**.

------------------------------------------------------------------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)

# Install packages if necessary (uncomment if needed):
install.packages(c("MatchIt", "sandwich", "lmtest" ,"marginaleffects"))
library(MatchIt)
library(sandwich)
library(lmtest)
library(marginaleffects)
```

## 1. Advanced Matching Methods

### 1(a) Load data and basic checks

```{r load-data}
# Load the data (make sure connors.rdata is in your working directory)
load("C:/Users/P094412/Documents/bcf-linear/connors.rdata")  # loads data as 'dat'

dim(dat)
summary(dat)
table(dat$W)
```

**Answer (a)**:\
Yes, the data are suitable for matching because there are more controls than treated units.

### 1(b) Simple 1:1 Nearest Neighbor Matching

```{r match1}
match1 <- matchit(
  W ~ sex + race + edu + age + aps + temp + meanbp + resp_rate + hrt_rate + paco2,
  data = dat,
  distance = "glm",    # logistic regression for propensity score
  method   = "nearest",# 1:1 NN matching
  ratio    = 1
)

summary(match1)
```

**Answer (b)**:\
No, balance criteria (\|SMD\| \< 0.1 and variance ratio near 1) are not fully met (e.g. `aps` is still imbalanced).

### 1(c) Improved Matching with Polynomial Terms

```{r match2}
# Fit a more flexible PS model with polynomial terms
psmod <- glm(
  W ~ sex + race + poly(edu, 2) + poly(age, 2) + poly(aps, 2) +
       poly(temp, 2) + poly(meanbp, 2) + poly(resp_rate, 2) +
       poly(hrt_rate, 2) + poly(paco2, 2),
  data   = dat,
  family = binomial()
)
ps <- predict(psmod, type = "response")

# Nearest neighbor with ratio=4, caliper=0.05, replacement=TRUE
match2 <- matchit(
  W ~ sex + race + edu + age + aps + temp + meanbp + resp_rate + hrt_rate + paco2,
  data     = dat,
  distance = ps,
  method   = "nearest",
  caliper  = 0.05,  # 0.05 of PS (on the SD scale)
  ratio    = 4,
  replace  = TRUE
)

summary(match2)
```

**Answer (c)**:\
Yes, we now achieve excellent balance.

### 1(d) Unmatched treated units

```{r unmatched}
m2_summary <- summary(match2)
m2_unmatched <- m2_summary$nn["Unmatched", "Treated"]
cat("Number of unmatched treated units:", m2_unmatched, "\n")
```

### 1(e) Compare `get_matches` vs. `match.data`

```{r match2-outputs}
# get_matches approach
gm <- get_matches(match2)
head(gm)

# match.data approach
md <- match.data(match2)
head(md)

cat("\nWeights in 'gm':\n")
print(table(gm$weights))

cat("\nWeights in 'md':\n")
print(table(md$weights))
```

**Answer (e)**:\
- With replacement matching, the same control can be matched multiple times.\
- `get_matches()` duplicates the control rows if they are used multiple times.\
- `match.data()` does not duplicate rows; instead, it inflates the control's weight to reflect multiple matches.

### 1(f) Causal Odds Ratio (ATT) for 30-day Death

```{r match2-effect}
fit <- glm(
  death30 ~ W,
  data    = gm,
  weights = gm$weights,
  family  = binomial()
)

# Cluster-robust standard errors (assuming 'id' and 'subclass' exist)
# If 'id' is not in dat, remove it from cluster
res   <- coeftest(fit, vcov. = vcovCL, cluster = ~id + subclass)
OR_est <- exp(res[2,1])
p_val  <- 2 * pnorm(abs(res[2,1]/res[2,2]), lower.tail=FALSE)

cat("Estimated OR for death30:", round(OR_est, 3), "\n")
cat("p-value:", round(p_val, 4), "\n")
```

**Answer (f)**:\
Estimate is about 1.07, not significant, unlike the original Connors et al. conclusion.

### 1(g) Full Matching

```{r match3}
match3 <- matchit(
  W ~ sex + race + edu + age + aps + temp + meanbp + resp_rate + hrt_rate + paco2,
  data     = dat,
  distance = ps,
  method   = "full",
  estimand = "ATT"
)

summary(match3)
```

**Answer (g)**:\
Balance is similarly excellent. Full matching typically uses *all* subjects by assigning them appropriate weights.

### 1(h) Effect with Full Matching

```{r match3-effect}
gm3 <- match.data(match3)
fit3 <- glm(
  death30 ~ W,
  data    = gm3,
  weights = gm3$weights,
  family  = binomial()
)

res3    <- coeftest(fit3, vcov. = vcovCL, cluster = ~subclass)
OR_est3 <- exp(res3[2,1])
cat("Full Matching OR:", round(OR_est3, 3), "\n")
```

**Answer (h)**:\
Very similar effect estimate, slightly bigger standard error.

### 1(i) ANCOVA-style regression after matching

```{r ancova}
msm3 <- glm(
  death30 ~ W + sex + race + edu + age + aps + temp + meanbp + resp_rate + hrt_rate,
  data    = gm3,
  weights = gm3$weights,
  family  = binomial()
)

avg_comparisons(
  msm3,
  variables  = "W",
  newdata    = subset(gm3, W == 1),   # for ATT
  comparison = "lnoravg",
  vcov       = ~subclass,
  wts        = "weights",
  transform  = exp
)
```

**Answer (i)**:\
Similar point estimate to the direct logistic approach, but potentially smaller residual bias and a more robust design-based inference.

------------------------------------------------------------------------

## 2. Double Robust Estimation (AIPW)

We use the following function for the risk difference version of AIPW:

```{r aipw-func}
ace_aipw <- function(y, A, w, m1, m0) {
  n  <- length(y)
  # influence components
  I1 <- A * w * y - w * (A - 1/w) * m1
  I0 <- (1 - A) * w * y - w * ((1 - A) - 1/w) * m0
  
  Ey1.est <- sum(I1) / n
  Ey0.est <- sum(I0) / n
  ace     <- Ey1.est - Ey0.est
  
  # approximate var (risk difference)
  I       <- I1 - I0
  var.ace <- (1/n^2) * sum((I - ace)^2)
  se.ace  <- sqrt(var.ace)
  
  c(Ey1.est = Ey1.est, Ey0.est = Ey0.est, ace = ace, `se(ace)` = se.ace)
}
```

### 2(a) & 2(b) Fit outcome & PS models with polynomials

```{r dr-models}
outcome_model <- death30 ~ sex + race + poly(edu,2) + poly(age,2) +
                            poly(aps,2) + poly(temp,2) + poly(meanbp,2) +
                            poly(resp_rate,2) + poly(hrt_rate,2) + poly(paco2,2)

mod0 <- glm(outcome_model, data=dat, family=binomial, subset=(W==0))
mod1 <- glm(outcome_model, data=dat, family=binomial, subset=(W==1))

psmod <- glm(
  W ~ sex + race + poly(edu,2) + poly(age,2) + poly(aps,2) +
       poly(temp,2) + poly(meanbp,2) + poly(resp_rate,2) +
       poly(hrt_rate,2) + poly(paco2,2),
  data   = dat,
  family = binomial()
)

summary(mod0)
summary(mod1)
summary(psmod)
```

### 2(c) & 2(d) Predictions + AIPW estimate (risk difference)

```{r dr-aipw}
py0 <- predict(mod0, newdata=dat, type="response")  # P(Y=1|X,A=0)
py1 <- predict(mod1, newdata=dat, type="response")  # P(Y=1|X,A=1)
ps  <- predict(psmod, newdata=dat, type="response") # P(A=1|X)

w <- ifelse(dat$W==1, 1/ps, 1/(1-ps))

res_aipw <- ace_aipw(y=dat$death30, A=dat$W, w=w, m1=py1, m0=py0)
res_aipw
```

**Answer**:\
Check the 95% CI with:

```{r dr-ci}
rd     <- res_aipw["ace"]
se_rd  <- res_aipw["se(ace)"]
ci_low <- rd - 1.96*se_rd
ci_up  <- rd + 1.96*se_rd

cat("AIPW Risk Difference:", round(rd,4),
    " with 95% CI = [", round(ci_low,4), ",", round(ci_up,4), "]\n")
```

Likely it is not significant at $\alpha=0.05$.

### 2(e) Advanced: AIPW for the Odds Ratio

```{r dr-aipw-or}
ace_aipw_or <- function(y, A, w, m1, m0) {
  n  <- length(y)
  I1 <- A * w * y - w * (A - 1/w) * m1
  I0 <- (1 - A) * w * y - w * ((1 - A) - 1/w) * m0
  
  Ey1.est <- sum(I1) / n
  Ey0.est <- sum(I0) / n
  
  # OR = [p1/(1-p1)] / [p0/(1-p0)]
  or_est   <- (Ey1.est / (1 - Ey1.est)) / (Ey0.est / (1 - Ey0.est))
  
  c(Ey1.est=Ey1.est, Ey0.est=Ey0.est, OR=or_est)
}

res_aipw_or <- ace_aipw_or(dat$death30, dat$W, w, py1, py0)
res_aipw_or
```

**Answer**:\
The AIPW OR is around \~1.06, which is not significant either.

------------------------------------------------------------------------

**End of Document**
